---
title: "Developing PGP calculations"
output: html_notebook
date: "2023-08-02"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

I simulated some data using `/Users/eriq/Documents/git-repos/MixedUpSlimSims/development/004-more-realistic-example-use.qmd`
which looks like cyclone creek, and I copied the results to this repo at `development/data/from-MUSS/004-sim-results.rds`
relative to this notebook.

Then in 004 in this repo I created the extended segments and saved them so I
have all this stuff to work with quickly.

So, I can read that all in now:
```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(MixedUpParents)

E <- read_rds("data/from-MUSS/extended_anc_segs.rds")

sim_results <- read_rds("data/from-MUSS/004-sim-results.rds")
V <- sim_results$var_snps

M <- sim_results$spp_diag %>%
  mutate(
    diag_spp = case_match(
      diag_spp,
      "WCT" ~ "A",
      "RBT" ~ "B",
      "YCT" ~ "C"
    )
  )

dir.create("dev-results/005", recursive = TRUE, showWarnings = FALSE)
```


See 004 for an exploration of the data and extended segments.  Here we just go immediately
into preparing for pairwise genotype probs:
```{r}
L <- prepare_for_gpp(V, M, E)


# Before going further, let's make sure we have the sampled individuals with their year and
# also their 0-based idx's

# get the total simulation pedigree
pedigree <- sim_results$metaQ %>%
  select(ind_time, sex, ped_id:ped_p2) %>%
  distinct() %>%
  mutate(across(starts_with("ped"), .fns = as.character))

# now pick out just the sampled ones that we have
samples <- L$ir_E$iKey %>%
  left_join(pedigree, by = join_by(indiv == ped_id))

# and now make vectors of the kids and the parents and grandparents that
# were sampled
kids <- samples %>% filter(ind_time == 0) %>% pull(indiv)
pars <- samples %>% filter(ind_time == 1) %>% pull(indiv)
gpars <- samples %>% filter(ind_time == 2) %>% pull(indiv)
```

Here I want to get some kids and their parents that were actually sampled.
```{r}
sampled_ped <- pedigree %>%
  mutate(
    ped_id = ifelse(ped_id %in% kids, ped_id, NA),
    ped_p1 = ifelse(ped_p1 %in% pars, ped_p1, NA),
    ped_p2 = ifelse(ped_p2 %in% pars, ped_p2, NA)
  ) %>%
  filter(!is.na(ped_id))

sampled_ped
```


Now, we are in a position to start developing and testing things.  Here we run `pairwise_genotype_probs()` with some of the individuals from above.

**BIG NOTE: I am running this having compiled pgp_rcpp.cpp up with
`const int WriteRcouts = 1;` So that it writes a bunch of stuff out.
To be honest.  I actually run this all in the regular terminal and then
I copy the output to another file to be processed with `scripts/munge-Rcouts.awk`**


Here is a kid with his two parents and then two randos
```{r}
TMP <- pairwise_genotype_probs(
  L = L,
  kid = "20014",
  par = c("19454", "19767", "18003", "18435")
)
```

I ran that and then put the output into: `005-stuff/pgp-stdouts/k-20014_p-19454-19767.stdout`.
Then I can process it in awk like this:
```sh
(base) /development/--% (main) pwd
/Users/eriq/Documents/git-repos/MixedUpParents/development

BASE="k-20014_p-19454-19767_u-18003-180345"
pbpaste > 005-stuff/pgp-stdouts/$BASE.stdout
awk -f ../scripts/munge-Rcouts.awk 005-stuff/pgp-stdouts/$BASE.stdout
mv P_unrel.txt 005-stuff/pgp-stdouts-awked/$BASE.unrel
mv P_parental.txt  005-stuff/pgp-stdouts-awked/$BASE.parental

```

Now I can read those in and analyze them.

## Reading in the trio and analyzing

```{r}
ppar <- read_table("005-stuff/pgp-stdouts-awked/k-20014_p-19454-19767_u-18003-180345.parental")
punrel <- read_table("005-stuff/pgp-stdouts-awked/k-20014_p-19454-19767_u-18003-180345.unrel")
```

Let's just go ahead and compare the genotype probs that we see for each of these.
Note that that these are not heavily admixed so their are not a whole lot of
segged pieces out of the parents, so let's go ahead and filter things down to
spots where the
```{r}
punrel %>% count(anc_p, anc_k)
```

So, quick check of the variable markers in the 6 6 sections.
```{r}


unrel_var_66 <- punrel %>% select(kIdx, pIdx, anc_k, anc_p, midx, gp, gk, isDiag, geno_prob) %>% filter(anc_k == 6, anc_p == 6, isDiag == 0)
ppar_var_66 <- ppar %>% select(kIdx, pIdx, anc_k, anc_p, midx, gp, gk, isDiag, geno_prob) %>% filter(anc_k == 6, anc_p == 6, isDiag == 0)

comp_up <- left_join(
  unrel_var_66,
  ppar_var_66,
  by = join_by(kIdx, pIdx, anc_k, anc_p, midx, gp, gk),
  suffix = c(".u", ".p")
)

# and make a plot
ggplot(comp_up, aes(x = geno_prob.u, y = geno_prob.p)) +
  geom_point() +
  facet_grid(gp + gk ~ pIdx) +
  geom_abline(intercept = 0, slope = 1)
```

OK, count those cases up
```{r}
comp_up %>%
  count(kIdx, pIdx, gp, gk) %>%
  mutate(Geno = str_c("Par-", gp, "  Kid-", gk)) %>%
  ggplot(aes(x = Geno, y = n, fill = factor(pIdx))) +
  geom_col() +
  theme(axis.text.x = element_text(angle = 90))
```

OK.  That looks pretty reasonable...now.  In other words I tracked down a few bugs and it
seems to be working.  Let's look at the distribution of the Logls...
```{r}
ggplot(punrel, aes(x = log(geno_prob))) +
  geom_histogram() +
  facet_grid(isDiag ~ gk)
```

Now, look at it for parental probs.  First for the two actual parents:
```{r}
ggplot(ppar %>% filter(pIdx %in% c(1407, 1547)), aes(x = log(geno_prob))) +
  geom_histogram() +
  facet_grid(isDiag ~ gk)
```


And then for the two non parents.
```{r}
ggplot(ppar %>% filter(!(pIdx %in% c(1407, 1547))), aes(x = log(geno_prob))) +
  geom_histogram() +
  facet_grid(isDiag ~ gk)
```

The really cool thing here is that we are getting the correct results for the actual LogLs, too.
In other words, the parents have much high LogLRatios.


## Checking on all the individuals 

I think it is time to recompile that thing with `const int WriteRcouts = 0;` and
run this on a lot of indivs:
```{r}
all_comps <- lapply(kids[1:100], function(k) {
  pairwise_genotype_probs(
  L = L,
  kid = k,
  par = pars
) %>%
  mutate(logl_ratio = probKidParental - probKidUnrel)
})


check <- all_comps %>%
  bind_rows() %>%
  arrange(desc(logl_ratio)) %>%
  left_join(L$ir_E$iKey, by = join_by(kIdx == idx)) %>%
  rename(kidsample = indiv) %>%
  left_join(L$ir_E$iKey, by = join_by(pIdx == idx)) %>%
  rename(parsample = indiv) %>%
  left_join(pedigree, by = join_by(kidsample == ped_id))

ggplot(check, aes(x = logl_ratio)) + geom_histogram() + facet_wrap(~logl_ratio > 0, scales = "free")
```
