---
title: "Parentage ROC Explanation Figures"
output: html_notebook
---

One of the referees noted that the parentage ROC curves are a helpful advance
in their own right and wanted to have something that walked the reader through
how they are constructed.  That seems like a fine idea. 

Here is how we will do it:  10 candidate offspring, and a separate set of
20 candidate parents.  We will shoot for 2 offspring that have both parents
correctly identified.  1 offspring where both parents are present, but only one
is correctly identified (the other "unrelated" pair has higher PO/U likelihood).
3 offspring with one parent in the data set, and they are all correctly identified.
1 offspring with one parent in the data set that is not correctly identified,
and 3 offspring with no parents in the data set.  


We will label the offspring 0-9.  The candidate parents will be labeled 10-29.
We will set up an artificial situation (that works well for reflecting logl-ratio
values by color, but does not really reflect the variance and expected values of different
categories well) where unrelateds have logl values with a mean around -20, and parentals
have logl values with a mean of around 10, but that some non-parental pairs will
have some a pretty wide variance, including some high values, which I might even put in there
by hand.

First generate and look at the all the logl values.
```{r}
library(tidyverse)

# here is a decent distro for true PO pairs
rgamma(2000, shape = 8, scale = 1/1) %>% hist()
```

```{r}
# here is a decent distro for true unrelateds
(rgamma(2000, shape = 8, scale = 1/1)- 15) %>% hist()
```

So, let's just simulated those:
```{r}
set.seed(5)
unrel_mat <- (rgamma(200, shape = 8, scale = 1/1) - 12) %>%
  matrix(nrow = 10) %>%
  apply(1, function(x) rev(sort(x))) %>%
  t()

po_mat <- (rgamma(20, shape = 8, scale = 1/1) - 2) %>%
  matrix(nrow = 10) %>%
  apply(1, function(x) rev(sort(x))) %>%
  t()

cbind(po_mat, unrel_mat[,1:2])
```
So 1 and 2 will be the two with both parents.  (0 and 1),  4 will be the one with both parents
but one is incorrectly identified (2).  3, 5, 6 will be the three with one parent and it
is correctly identified (3, 4, 5).  9 will the the one with one parent and it is incorrectly
identified (6).  And then the remaining ones: 7, 8, 10, will be the ones with no parents
in there at all.  In fact, I won't make the base-0 indices in order like that, I will just
keep them in the order they are in (seems more realistic and not that hard to describe).
So, let's implement all that.

First, do up the PO_mat into a tib.
```{r}
colnames(po_mat) <- c("p1", "p2")
po_mat[c(3,5,6), ] [,1] <- NA
po_mat[c(7,8,10),] <- NA
po_mat[9,1] <- NA
po_mat[3,2] <- 3.1345   # set this loe
po_mat[1,1] <- NA   # Drop that one parent


PO_logls_tib <- as_tibble(po_mat) %>%
  mutate(off_idx = 0:9) %>%
  pivot_longer(-off_idx) %>%
  mutate(relat = "PO") %>%
  filter(!is.na(value)) %>%
  select(-name)
```

Then, do up the unrel ones, including doctoring a few of the numbers that need doctoring.
```{r}
unrel_mat[9,1] <- 3.8223
unrel_mat[9,2] <- 3.73248  # this means that the true parent is missed all together for this one
unrel_mat[3, 1] <- 3.22395
colnames(unrel_mat) <- paste0("u", 1:ncol(unrel_mat))

U_logls_tib <- as_tibble(unrel_mat) %>%
  mutate(off_idx = 0:9) %>%
  pivot_longer(-off_idx) %>%
  mutate(relat = "U") %>%
  select(-name)
```

Now, that we have those, we can combine them, and retain the top 20 parents for offspring, and
we can just simulate the IDs of the parents into the offspring slots 
```{r}
full_tib <- bind_rows(PO_logls_tib, U_logls_tib) %>%
  arrange(off_idx, desc(value)) %>%
  group_by(off_idx) %>%
  slice(1:20) %>%
  mutate(rank = 1:20) %>%
  ungroup()
```

Now, we can check on the number of parents each has, and their ranks:
```{r}
full_tib %>%
  group_by(off_idx) %>%
  filter(relat == "PO")
```

I don't like having the parent-pair-present ones all in the beginning, so I will swap 0 and 6 and 1 and 9.
And then I will pass the IDs in for the candidate parents as single letters.  
```{r}
full_tib2 <- full_tib %>%
  mutate(
    off_idx = case_when(
      off_idx == 0 ~ 6L,
      off_idx == 6 ~ 0L,
      off_idx == 1 ~ 9L,
      off_idx == 9 ~ 1L,
      TRUE ~ off_idx
    )
  ) %>%
  mutate(bigx = ifelse(rank > 2, rank + 0.5, rank)) %>%
  group_by(off_idx) %>%
  mutate(par_id = sample(letters[1:20])) %>%
  ungroup() %>%
  mutate(
    text_color = ifelse(value < 3.5, "white", "black")
  )

full_tib2 %>%
  group_by(off_idx) %>%
  filter(relat == "PO")
```

So, now we can make a plot of little circles showing the logl-ratio values.
```{r}
POPs <- full_tib2 %>%
  filter(relat == "PO")

roc_fill_limits <- range(full_tib2$value, na.rm = TRUE)
roc_fill_scale <- function() {
  scale_fill_viridis_c(option = "plasma", limits = roc_fill_limits, name = "log LR or\nScore")
}
roc_output_dir <- "outputs/007"
dir.create(roc_output_dir, recursive = TRUE, showWarnings = FALSE)

OffIDs <- tibble(
  y = 0:9
) %>%
  mutate(
    id_text = as.character(rev(y)),
    x = -.3
  )

g1 <- ggplot(full_tib2, aes(x = bigx, y = off_idx)) +
  geom_point(aes(fill = value), shape = 21, size = 7) +
  geom_point(data = POPs, shape = 21, size = 9.5, fill = NA) +
  geom_text(aes(label = par_id, color = text_color)) +
  geom_point(data = OffIDs, aes(x = x, y = y), shape = 22, size = 7) +
  geom_text(data = OffIDs, aes(x = x, y = y, label = id_text, color = "black")) +
  roc_fill_scale() +
  theme_bw() +
  theme(
    panel.grid = element_blank(),
    axis.ticks = element_blank(),
    axis.text = element_blank()
  ) +
  scale_colour_manual(values = c(white = "white", black = "black"), guide = "none") +
  coord_cartesian(xlim = c(-0.9, max(full_tib2$bigx) + 0.9), ylim = c(-0.7, 9.7), clip = "off") +
  xlab("Candidate Parent Rank (Within Each Offspring)") +
  ylab("Candidate Offspring Index") +
  geom_vline(xintercept = 2.78, linewidth = 1.1)

ggsave(file.path(roc_output_dir, "roc-explanation-panel-1.pdf"), g1, width = 9.5, height = 4.0)
g1
```


OK.  That is a good mockup of the first panel.  Now we need to sort out the top
two candidate parents from each, and display that.  We put the PO Pairs with the
parent candidate up and the offspring down and to the right.  

```{r}
PO_top_two <- full_tib2 %>%
  filter(rank <= 2) %>%
  arrange(desc(value)) %>%
  mutate(
    dyad_rank = 1:n(),
    offs_id = as.character(-(off_idx - 9))
  )

g2 <- ggplot(PO_top_two) +
  geom_segment(aes(x = dyad_rank, y = 1, xend = dyad_rank + 1, yend = 0), colour = "gray") +
  geom_point(aes(x = dyad_rank,  fill = value), y = 1, shape = 21, size = 7) +
  geom_point(data = PO_top_two %>% filter(relat == "PO"), aes(x = dyad_rank), y = 1, shape = 21, size = 9.5, fill = NA) +
  geom_text(aes(x = dyad_rank, label = par_id, color = text_color), y = 1) + 
  geom_point(aes(x = dyad_rank + 1),  y = 0, shape = 22, size = 7, fill = "white") +
  geom_text(aes(x = dyad_rank + 1, label = offs_id), y = 0, color = "black") +
  roc_fill_scale() +
  scale_colour_manual(values = c(white = "white", black = "black"), guide = "none") +
  theme_bw() +
  theme(
    panel.grid = element_blank(),
    axis.ticks = element_blank(),
    axis.text = element_blank()
  ) +
  coord_cartesian(xlim = c(0.25, max(PO_top_two$dyad_rank) + 1.75), ylim = c(-0.6, 1.6), clip = "off") +
  ylab("") +
  xlab("Ranking of Candidate Parent Log-likelihoods (or other Scores)")

ggsave(file.path(roc_output_dir, "roc-explanation-panel-2.pdf"), g2, width = 9.5, height = 1.8)
g2
```


And now all we need to do is plot this thing in Parentage ROC curve format.
```{r}
MaxNP <- sum(full_tib2$relat == "PO")
MaxAllAssign <- 2 * n_distinct(full_tib2$off_idx)
POTT2 <- PO_top_two %>%
  mutate(
    cumul_right = cumsum(relat == "PO"),
    cumul_wrong = cumsum(relat != "PO"),
    xval_circ = cumul_wrong / MaxAllAssign,
    yval_circ = cumul_right / MaxNP,
    xval_sq = xval_circ + 1.5 / MaxAllAssign,
    yval_sq = yval_circ - 0.8 / MaxNP
  )
 

g3 <- ggplot(POTT2) +
  geom_line(data = bind_rows(tibble(xval_circ = 0, yval_circ = 0), POTT2), aes(x = xval_circ, y = yval_circ), linewidth = 1.5) +
  geom_segment(aes(x = xval_circ, xend = xval_sq, y = yval_circ, yend = yval_sq), colour = "gray") +
  geom_point(aes(x = xval_circ, y = yval_circ,  fill = value), shape = 21, size = 7) +
  geom_point(data = POTT2 %>% filter(relat == "PO"), aes(x = xval_circ, y = yval_circ), shape = 21, size = 9.5, fill = NA) +
  geom_text(aes(x = xval_circ, y = yval_circ, label = par_id, color = text_color)) +
  ylim(0, 1) +
  xlim(0, 1) +
  geom_point(aes(x = xval_sq,  y = yval_sq), shape = 22, size = 7, fill = "white") +
  geom_text(aes(x = xval_sq, y = yval_sq, label = offs_id), color = "black") +
  scale_fill_viridis_c(option = "plasma", name = "log LR or\nScore") +
  scale_colour_manual(values = c(white = "white", black = "black"), guide = "none") +
  theme_bw() +
  xlab("Fraction of Possible Offspring-Parent Assignments That are Wrong") +
  ylab("Fraction of All True Parent-Offspring Pairs Identified Correctly")

ggsave(file.path(roc_output_dir, "roc-explanation-panel-3.pdf"), g3, width = 9.5, height = 4.6)
g3
```
