---
title: "The Nascent Workflow"
output: html_notebook
---

This is just a notebook to remind me how to get to the bleeding edge of
development with the small test data set so that I can pick up working
on things easily.

The example to plot_ancestral_segs summarizes it pretty well:

**Hey! You can't do this fast from within RStudio because it is unstable with
respect to parallel processing with the future package**

```{r}
library(tidyverse)
library(MixedUpParents)
# get test data set as M
M <- diag_markers_10_fish

# find the ancestral-n-segments
D <- ancestral_n_segments(M, unique(M$chrom))


#' find the extended segments:
#' set the number of cores:
future::plan(future::multicore, workers = 8)  # 8 cores on my laptop


# extend those ancestral segments
E <- extend_ancestral_segments_3(D, c("WCT", "RBT", "YCT"))


# if you want to plot things you can:
#' # g2 <- plot_ancestral_segs(D, M, E = E)
#'
#' #' # with many individuals, it is recommended to plot this large:
#' # like: ggsave(g2, filename = "E-gplot.pdf", width = 30, height = 40)
```

Then, continuing, I can turn the E tibble into something that is suitable for
intersecting intervals in RCpp:
```{r}
Eir <- integer_representation_EAS(E)
```

And with a data structure like Eir, I can intersect the intervals on the different
chromosomes between whatever pairs of indivs I want to.  This is done with the Rcpp
function, intersect_ancestry_intervals_rcpp().  
```{r}
boing <- intersect_ancestry_intervals_rcpp(
  grp1 = 0:3,
  grp2 = 6:9,
  nC = Eir$nChrom,
  V1 = Eir$IntervalsMatrix,
  V2 = Eir$IntervalsMatrix,
  X1 = Eir$IndexMatrix,
  X2 = Eir$IndexMatrix,
  nv1 = Eir$nIntvlsVec,
  nv2 = Eir$nIntvlsVec
) 
```

OK! The above function seems to be working, but there are some things that I need to do:

1. Merge adjacent intervals in `extend_ancestral_segments_3()`.  I think there is a function for
that.  The issue can be seen in the figure for the segments for `OcLang16_185` on some of the chromosomes
where there was one marker that had an aberrant call.  Basically, if two segments with the same ancestry
are within 10 bp of one another they should be merged, I think.
