---
title: "Pre-Processing data from Cyclone Creek"
output: html_document
---

Here we do a data set of about 4000 fish.

```{r}
library(tidyverse)
library(MixedUpParents)
library(vroom)

# read in all the files:
cyc_ck_files <- dir("BigData/CycloneCreek/allele_counts", full.names = TRUE)

all_files <- vroom(cyc_ck_files, id = "path")

```

There is a problem with a few loci that have more than two alleles:
```{r}
problems(all_files)
```

Let's count those up:
```{r}
problems(all_files) %>%
  count(row, expected, actual)
```

So, let's just read everything as characters, pivot this stuff, and then NA-call the sites
that have a `/`.
```{r}
all_files_chr <- vroom(
  cyc_ck_files,
  col_types = cols(.default = "character"),
  id = "path"
  ) %>%
  pivot_longer(
    cols = -(path:Pos),
    names_to = "fish",
    values_to = "geno"
  ) %>%
  extract(path, into = "marker_set", "^Big.*allele_counts/(.*)_allele_counts.csv")
```


Count up the different genotypes we have seen here:
```{r}
all_files_chr %>%
  count(geno)
```
So, there are clearly some that have more than the two main alleles.
We will simply no-call those, as those genotype calls are a tiny subset of the others, see:
```{r}
all_files_chr %>%
  count(geno) %>%
  mutate(
    geno_type = case_when(
      geno %in% c("0", "1", "2") ~ "is012",
      is.na(geno) ~ "isNA",
      str_detect(geno, "/") ~ "hasSlash",
      TRUE ~ NA
    )
  ) %>%
  group_by(geno_type) %>% 
  summarise(tot_n = sum(n)) %>%
  mutate(
    fract = tot_n / sum(tot_n)
  )
```

That is a crap-ton of missing data. It would make little difference to chuck out the hasSlashes as NAs, too.

```{r}
genos <- all_files_chr %>%
  mutate(
    geno = as.integer(ifelse(str_detect(geno, "/"), NA, geno)),
  ) %>%
  rename(chrom = Chr, pos = Pos) %>%
  mutate(pos = as.integer(pos))
  
```

## Check for and remove loci that appear more than once

```{r}
duped_positions <- genos %>%
  count(chrom, pos, fish) %>%
  filter(n > 1) %>%
  count(chrom, pos)
```
And see where those appear:
```{r}
count_dupes <- genos %>%
  semi_join(duped_positions, by = c("chrom", "pos")) %>%
  count(marker_set, chrom, pos)

count_dupes
```

So, we can count up how many times these occur in each data set and then
pivot that to see the overlaps:
```{r, rows.print = 40}
dup_comp <- count_dupes %>%
  select(-n) %>%
  count(chrom, pos, marker_set) %>%
  pivot_wider(
    names_from = marker_set,
    values_from = n,
    values_fill = 0L
  )
dup_comp
```

I find it concerning when something is listed as both a WCT_diag and also a
WCT_var.  Same with RBT.  So, really the only ones of these that I feel like
preserving would be the ones that are considered both WCT_var and RBT_var, but
I will drop those from RBT_var.  All others must go.  So, we can do some
anti-joining here:
```{r}
drop_dup_from_RBT <- dup_comp %>%
  filter(WCT_var == 1 & RBT_var == 1 & WCT_diag == 0 & RBT_diag == 0) %>%
  mutate(marker_set = "RBT_diag", .before = chrom)
```

That is only 5 markers. Why even bother.  I am going to discard all of these
duplicates.
```{r}
genos_no_dup <- genos %>%
  anti_join(dup_comp, by = c("chrom", "pos"))
```

Let's see if that was successful:
```{r}
genos_no_dup %>%
  count(chrom, pos, fish) %>%
  filter(n > 1) %>%
  count(chrom, pos)
```
Yay! The offending loci in multiple marker sets have been eliminated.


## total fraction of missing sites within individual across all marker sets

```{r}
genos_no_dup %>%
  group_by(fish) %>%
  summarise(fract_missing = mean(is.na(geno))) %>%
  ggplot(aes(x = fract_missing)) +
  geom_histogram()
```

Now, I am curious what that looks like in each of the different marker sets:
```{r}
fmiss_by_set <- genos_no_dup %>%
  group_by(marker_set, fish) %>%
  summarise(fract_missing = mean(is.na(geno)))

ggplot(fmiss_by_set, aes(x = fract_missing, fill = marker_set)) +
  geom_histogram() +
  facet_wrap(~ marker_set, ncol = 3)
```

I think that we should consider a miss fract cutoff of 0.3, because for individuals
missing 30% of their loci, you expect that a fraction greater than 50% will still
be non-missing between them.  I am going to impose this for now and we can
follow up on it more, later.


But I am also rather worried that some markers might be totally missing in all
backgrounds, though that would not make sense, I think.  

How about for different loci?
```{r}
genos_no_dup %>%
  group_by(chrom, pos) %>%
  summarise(fract_missing = mean(is.na(geno))) %>%
  ggplot(aes(x = fract_missing)) +
  geom_histogram()
```


When I look at that, I think I am inclined to discard markers that are missing
in more than 70% of the fish.  So, combining the two filters (loci and then indivs)
that I would like to do, we have something like this:




```{r}
genos_filt <- genos_no_dup %>%
  group_by(chrom, pos) %>%
  filter(mean(is.na(geno)) < 0.7) %>%
  group_by(fish) %>%
  filter(mean(is.na(geno)) < 0.3) %>%
  ungroup()
```

Now, let's see what that looks like in terms of numbers:
```{r}
genos_filt %>%
  count(chrom, pos) %>%
  count(n)
```

That is 2464 loci, and 3131 fish, total.

Unfiltered we had:
```{r}
genos_no_dup %>%
  count(chrom, pos) %>%
  count(n)
```
So, unfiltered was 2580 markers in 4178 fish.  

So, our missing-fraction-in-individuals filter has tossed about 1,050
fish.  But our locus filter did not toss too many loci.  We can revisit
that, but I am not eager to start doing things with only 25% of the original
markers in any particular pairwise comparison, which is what we would have
if we accepted fish with 50% missing data.

So, what do the number of markers in each of the data sets look like?
```{r}
genos_filt %>%
  group_by(marker_set) %>%
  summarise(
    num_fish = n_distinct(fish),
    num_loci = n_distinct(chrom, pos)
  )
```
Those numbers should work out pretty well, especially for individuals with
mostly WCT ancestry.  We might be able to loosen up on the missing data cutoffs
eventually.  We do seem to not have a great many WCT_diag markers, but that
might be OK.  


Let's also look at the patterns of missingness in the different sets of markers, again:
```{r}
genos_filt %>%
  group_by(marker_set, fish) %>%
  summarise(fract_missing = mean(is.na(geno))) %>%
  ggplot(aes(x = fract_missing, fill = marker_set)) +
  geom_histogram() +
  facet_wrap(~ marker_set, ncol = 3)
```

And let us compare that to what we were working with before:
```{r}
diag_markers_10_fish %>%
  group_by(diag_spp) %>%
  summarise(
    num_loci = n_distinct(chrom, pos)
  )
```

So, it is close.

OK,  we now have the cyclone creek data set in reasonable shape to start some testing.

```{r}
genos_filt_split <- genos_filt %>%
  separate(marker_set, into = c("species", "what"), sep = "_") %>%
  rename(
    indiv = fish,
    n = geno
  )

diags <- genos_filt_split %>%
  filter(what == "diag") %>%
  rename(diag_spp = species) %>%
  select(-what) %>%
  arrange(indiv, chrom, pos, diag_spp)

var_sites <- genos_filt_split %>%
  filter(what == "var") %>%
  rename(var_spp = species) %>%
  select(-what) %>%
  arrange(indiv, chrom, pos, var_spp)

dir.create("BigData/CycloneCreek/Processed", showWarnings = FALSE)

write_rds(
  diags,
  file = "BigData/CycloneCreek/Processed/spp_diagnostics.rds",
  compress = "xz"
)

write_rds(
  var_sites,
  file = "BigData/CycloneCreek/Processed/var_sites.rds",
  compress = "xz"
)
```
