---
title: "Example run of MixedUpParents"
output: rmarkdown::html_vignette
header-includes:
  - \usepackage{mathrsfs}
vignette: >
  %\VignetteIndexEntry{Example run of MixedUpParents}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(tidyverse)
library(MixedUpParents)

threads <- 8
```

This vignette walks through a small example analysis with `MixedUpParents`. The
example data set was generated by the simulation workflow in
`MixedUpParents-sims-and-analyses`. It starts from a SLiM simulation of an
admixed two-population pedigree, then applies subsampling, genotyping error, and
missingness to produce a realistic `MixedUpParents` input object.

Because the object is simulated, it also retains the true pedigree and
relationships. Those simulation-truth objects are useful for checking the
results below, but they are not required for running `MixedUpParents` on real
data.

```{r load-input-data}
T2M_path <- system.file("extdata", "tweaked2mup.rds", package = "MixedUpParents")
input_data <- read_rds(T2M_path)

names(input_data)
```

The required genotype inputs are long-format tibbles. `variable_snps` contains
genotypes at markers that are variable within ancestry backgrounds, with one row
per individual-marker observation and columns `chrom`, `pos`, `indiv`, and `n`.
`spp_diag_snps` has the same structure, but also includes `diag_spp`, the
ancestry for which the allele is diagnostic. In both tables, `n` is the observed
allele dosage, coded as 0, 1, 2, or `NA` when the genotype is missing.

```{r summarize-required-inputs}
input_data$variable_snps %>%
  summarise(
    rows = n(),
    individuals = n_distinct(indiv),
    chromosomes = n_distinct(chrom),
    markers = n_distinct(paste(chrom, pos)),
    missing_genotypes = sum(is.na(n))
  )

input_data$spp_diag_snps %>%
  summarise(
    rows = n(),
    individuals = n_distinct(indiv),
    chromosomes = n_distinct(chrom),
    markers = n_distinct(paste(chrom, pos)),
    missing_genotypes = sum(is.na(n))
  )
```

The diagnostic markers are labeled by the ancestry for which the alternate allele
is diagnostic. In this simulated example there are three diagnostic ancestries,
although the simulated admixed pedigree was generated from two focal populations.

```{r diagnostic-marker-counts}
input_data$spp_diag_snps %>%
  distinct(diag_spp, chrom, pos) %>%
  count(diag_spp, name = "diagnostic_markers")
```

The remaining elements of `input_data` document the simulation or let us check
the method after the fact. `sampled_pedQ` contains the true pedigree, sampling
generation, population, and true admixture fraction. `pairwise_relats` contains
the true relationship class for pairs of sampled individuals. Neither table is
needed in an empirical analysis.

```{r simulation-truth-preview}
input_data$sampled_pedQ %>%
  select(ped_id, ped_p1, ped_p2, ind_pop, ind_time, anc_pop, admix_fract) %>%
  slice_head(n = 6)
```

For the rest of the vignette we use the shorter name `TW`, matching the
development scripts that produced this example.

```{r name-example-data}
TW <- input_data
```

## Infer ancestry segments

The diagnostic markers are first converted into ancestry copy-number segments.
`ancestral_n_segments()` finds stretches of chromosome over which each
individual has the same number of copies of each diagnostic ancestry.

In the notation of the paper, these segments are the empirical version of the
ancestry tracts \(\mathscr{T}_w\) for individual \(w\). For a particular segment
\(s\), the ancestry state is written \(\mathscr{T}_w^{(s)}\).

This step, and the extension step that follows, take a little time, but not as
much as the calculating the pairwise likelihoods later.

```{r ancestral-n-segments}
AS <- ancestral_n_segments(
  D = TW$spp_diag_snps,
  chrom_levels = unique(TW$spp_diag_snps$chrom)
)

AS
```

The next step combines the ancestry-specific copy-number segments into full
ancestry states. For a diploid individual with three possible ancestries, the
possible states include values such as `A2B0C0`, `A1B1C0`, and `A0B2C0`.

These labels correspond to the paper's \(A_xB_yC_z\) notation for
\(\mathscr{T}_w^{(s)}\). For example, `A1B1C0` means that, in that segment, the
individual carries one copy of ancestry `A`, one copy of ancestry `B`, and no
copies of ancestry `C`.

Here the biological labels are ordered as `WCT`, `RBT`, and `YCT`. That ordering
defines the internal labels `A`, `B`, and `C`.

```{r extend-ancestry-segments}
species_levels <- c(TW$spp_names, "YCT")

E <- extend_ancestral_segments_using_bedtools(
  AS,
  species_levels = species_levels
)

E
```

Thus, the object `E` is the main bridge between the observed diagnostic marker
data and the ancestry-tract notation in the paper. Each row gives one segment
\(s\) for one individual, with its chromosome, start and stop positions, and
copy-number state \(\mathscr{T}_w^{(s)}\).

These extended segments are useful diagnostic objects in their own right. The
first plot summarizes the fraction of each individual's genome assigned to each
full ancestry state. The black line shows the admixture fraction for ancestry
`A`, which is `WCT` in this example.

```{r ancestry-fraction-plot, fig.width = 10, fig.height = 10, dev = "svg"}
ancestry_barplot(E)
```

The second plot shows the same ancestry states laid out along the genome. This
is often a good way to spot unusual individuals or regions where diagnostic
markers are sparse.

```{r ancestry-segment-plot, fig.width = 10, fig.height = 30, dev = "svg"}
ancestry_barplot(E, plot_type = "segments")
```

## Prepare genotype probability inputs

`pairwise_genotype_probs()` uses both diagnostic and variable markers. Before we
can call it, `prepare_for_gpp()` estimates allele frequencies in each ancestry
background, estimates individual admixture fractions from the extended ancestry
segments, and builds integer-indexed data structures for the faster C++ code.

In the paper's notation, the diagnostic and variable markers in segment \(s\)
are the sets \(D^{(s)}\) and \(V^{(s)}\). The corresponding allele-frequency
matrices are \(\mathbf{\hat{P}}_D^{(s)}\) and \(\mathbf{\hat{P}}_V^{(s)}\), and
the estimated admixture fractions for an individual are
\(\tilde{\mathbf{Q}}_w\).

One detail is important: after the extended ancestry segments have been encoded
as `A`, `B`, and `C`, the diagnostic marker labels must use those same internal
labels. The conversion below preserves the same ancestry ordering used above.

```{r relabel-diagnostic-markers}
new_diag <- TW$spp_diag_snps %>%
  mutate(
    ancestry_index = as.integer(factor(diag_spp, levels = species_levels)),
    diag_spp = c("A", "B", "C")[ancestry_index]
  ) %>%
  select(diag_spp, chrom, pos, indiv, n)

new_diag %>%
  distinct(diag_spp, chrom, pos) %>%
  count(diag_spp, name = "diagnostic_markers")
```

Now we can prepare the object that will be passed to
`pairwise_genotype_probs()`.

```{r prepare-for-gpp}
PRP <- prepare_for_gpp(
  V = TW$variable_snps,
  M = new_diag,
  E = E,
  diag_err_rate = 0.005
)

names(PRP)
```

For example, `PRP$AF` contains the estimated allele frequencies for both
variable and diagnostic markers in the ancestry backgrounds, and `PRP$AD`
contains the estimated admixture fractions for each individual.

These correspond to the paper's \(\mathbf{\hat{P}}_D\),
\(\mathbf{\hat{P}}_V\), and \(\tilde{\mathbf{Q}}\) terms. The package stores
them in a form that can be indexed efficiently when comparing a candidate
offspring \(k\) to a candidate parent \(p\).

```{r prepared-object-preview}
PRP$AF %>%
  slice_head(n = 6)

PRP$AD %>%
  slice_head(n = 6)
```

## Compare candidate parent-offspring pairs

In an empirical analysis, the candidate offspring and candidate parents would
usually come from sampling dates, ages, cohorts, or other metadata. In this
simulated data set, we use the true simulation generation to construct the
candidate sets. Final-generation individuals are treated as candidate offspring,
and all sampled individuals are treated as candidate parents.

The pedigree table has two rows per individual, one for each ancestry-specific
admixture fraction. We keep one row per individual for metadata joins.

```{r choose-candidates}
qped <- TW$sampled_pedQ %>%
  filter(anc_pop == 1)

kids <- qped %>%
  filter(ind_time == 0) %>%
  pull(ped_id) %>%
  unique()

all_candi <- qped %>%
  pull(ped_id) %>%
  unique()

tibble(
  candidate_offspring = length(kids),
  candidate_parents = length(all_candi)
)
```

We run the pairwise probability calculation for all final-generation offspring.
This chunk is computationally intensive; the `threads` object set in the setup
chunk controls the number of parallel workers used by `parallel::mclapply()`.

For each ordered pair, `pairwise_genotype_probs()` intersects the ancestry
segments of the candidate parent and offspring. This gives the segment set
\(s \in p \cap k\) used in Equations 1 and 2 of the paper: Equation 1
(`eq:L0`) defines \(L_0(k \mid p)\), the conditional probability of the
offspring data under the unrelated hypothesis, and Equation 2 (`eq:L1`) defines
\(L_1(k \mid p)\), the corresponding probability under the parent-offspring
hypothesis.

```{r run-mixedupparents}
run_mup <- function(kids, pars, L) {
  parallel::mclapply(
    kids,
    function(kid) {
      candi <- setdiff(pars, kid)

      pairwise_genotype_probs(
        L = L,
        kid = kid,
        par = candi
      )
    },
    mc.cores = threads
  ) %>%
    bind_rows()
}

if(system2("hostname", stdout = TRUE)[1] == "Erics-BGP-Laptop.local") {
  # For local vignette rendering while editing, eric uses the precomputed result.
  # This code should not run on anyone else's computer.  Instead, it will run
  # run_mup() for a long time!
  logls <- read_rds("../mup_logls.rds")
} else {
  logls <- run_mup(kids, all_candi, PRP) %>%
    mutate(logl_ratio = probKidParental - probKidUnrel) %>%
    arrange(desc(logl_ratio))
}


logls %>%
  select(kid_id, par_id, probKidUnrel, probKidParental, logl_ratio,
         nonMissingDiag, nonMissingVar) %>%
  slice_head(n = 10)
```

The column `logl_ratio` is the log-likelihood under the parent-offspring model
minus the log-likelihood under the unrelated model. Larger values indicate
stronger support for a candidate parent-offspring relationship. In the notation
of the paper,
\[
\Lambda_{k,p}^{\mathrm{PO/U}} =
\log\left(\frac{L_1(k \mid p)}{L_0(k \mid p)}\right).
\]
The package returns log probabilities, so the code calculation
`probKidParental - probKidUnrel` is exactly this log-likelihood ratio.

```{r plot-log-likelihood-ratios, fig.width = 7, fig.height = 4}
ggplot(logls, aes(x = logl_ratio)) +
  geom_histogram(bins = 60) +
  geom_vline(xintercept = 0, colour = "red", linewidth = 0.25) +
  xlab("log likelihood ratio: parent-offspring minus unrelated") +
  ylab("candidate pairs") +
  theme_bw()
```

## Check against the simulated pedigree

The package does not require a pedigree. Here, though, the simulation truth lets
us check whether the highest-scoring candidate pairs are real relatives.

First we record whether each offspring's true parents were included among the
sampled candidates.

```{r parents-in-sample}
pars_in_samples <- qped %>%
  filter(ped_id %in% kids) %>%
  select(ped_id, ped_p1, ped_p2) %>%
  mutate(
    p1_in_sample = ped_p1 %in% all_candi,
    p2_in_sample = ped_p2 %in% all_candi,
    num_parents_in_sample = p1_in_sample + p2_in_sample
  ) %>%
  mutate(
    across(c(ped_id, ped_p1, ped_p2), as.character)
  )

pars_in_samples
```

Next we attach simulation metadata and true pairwise relationship labels. This
join is only for interpretation of the simulated example; it is not part of the
required `MixedUpParents` workflow.

```{r join-simulation-truth}
meta <- qped %>%
  select(ped_id, ind_pop, ind_time, admix_fract) %>%
  rename(id = ped_id, pop = ind_pop, time = ind_time, q1 = admix_fract) %>%
  mutate(id = as.character(id))

kid_meta <- meta
names(kid_meta) <- str_c("kid", names(meta), sep = "_")

par_meta <- meta
names(par_meta) <- str_c("par", names(meta), sep = "_")

all_pairs <- logls %>%
  left_join(kid_meta, by = join_by(kid_id)) %>%
  left_join(par_meta, by = join_by(par_id)) %>%
  mutate(
    min_id = as.character(pmin(as.integer(kid_id), as.integer(par_id))),
    max_id = as.character(pmax(as.integer(kid_id), as.integer(par_id)))
  ) %>%
  left_join(
    TW$pairwise_relats %>%
      select(id_1, id_2, dom_relat, max_hit),
    by = join_by(min_id == id_1, max_id == id_2)
  ) %>%
  select(-min_id, -max_id) %>%
  mutate(
    dom_relat = replace_na(dom_relat, "U"),
    max_hit = replace_na(max_hit, 0L),
    rel_label = case_when(
      dom_relat == "A" ~ str_c("AN-", max_hit),
      dom_relat %in% c("Si", "FC") ~ str_c(dom_relat, "-", max_hit),
      TRUE ~ dom_relat
    ),
    po_fs_logl_ratio = probKidParental - probKidFullSib,
    po_fs_support = if_else(
      po_fs_logl_ratio > 0,
      "PO/FS > 0",
      "PO/FS <= 0"
    )
  )

all_pairs %>%
  select(kid_id, par_id, logl_ratio, po_fs_logl_ratio, rel_label,
         dom_relat, max_hit,
         kid_time, par_time, kid_q1, par_q1) %>%
  slice_head(n = 10)

plot_pairs <- all_pairs %>%
  filter(!(dom_relat %in% c("A", "Si", "FC") & max_hit > 2))
```

Because this is simulated data, we can also look at the likelihood-ratio
distributions by the true relationship class. The relationship labels come from
the pedigree simulation, not from `MixedUpParents`. Some relationship classes
are further subdivided by the `max_hit` column: for example, `Si-1` and `Si-2`
are sibling relationships with one or two major shared ancestors, corresponding
roughly to half siblings and full siblings. We use the same convention for `A`
and `FC`, but display the `A` class as `AN`, yielding labels such as `AN-1`,
`AN-2`, `FC-1`, and `FC-2`. The label `PO` denotes true parent-offspring pairs,
and `U` denotes pairs that do not appear in the simulated relationship table. A
few simulated relationships have `max_hit` greater than 2; we omit those rare
categories from the plots below.

```{r logl-ratio-by-relationship, fig.width = 8, fig.height = 6}
ggplot(plot_pairs, aes(x = logl_ratio)) +
  geom_histogram(bins = 80) +
  geom_vline(xintercept = 0, colour = "red", linewidth = 0.25) +
  facet_wrap(~ rel_label, scales = "free") +
  xlab("log likelihood ratio: parent-offspring minus unrelated") +
  ylab("candidate pairs") +
  theme_bw()
```

The C++ calculation also returns the log likelihood under a full-sibling model.
Rather than plotting that quantity for every relationship class, it is most
useful to focus on the two classes that the parent-offspring and full-sibling
models are meant to distinguish. In the plot below, `Si-2` denotes simulated
full siblings, `Si-1` denotes half siblings, `AN-2` denotes one of the close
ancestral relationship classes, and `PO` denotes simulated parent-offspring
pairs. The x-axis is the parent-offspring vs. unrelated log-likelihood ratio,
\(\log(L_1(k \mid p) / L_0(k \mid p))\). The y-axis is the parent-offspring vs.
full-sibling log-likelihood ratio,
\(\log(L_1(k \mid p) / L_\mathrm{FS}(k \mid p))\). A common screening rule is
to discard candidate parent-offspring pairs when the full-sibling model is more
likely than the parent-offspring model. In the paper's terms this is
\(\log(L_\mathrm{FS}/L_1) > 0\), which is equivalent to the plotted quantity
\(\log(L_1/L_\mathrm{FS}) < 0\). Thus, points below the horizontal zero line
would be treated as possible full-sibling pairs rather than retained as
parent-offspring calls.

```{r full-sib-vs-parental-logls, fig.width = 7, fig.height = 5}
all_pairs %>%
  filter(rel_label %in% c("Si-1", "Si-2", "AN-2", "PO")) %>%
  ggplot(aes(x = logl_ratio, y = po_fs_logl_ratio, colour = po_fs_support)) +
  geom_point(alpha = 0.3, size = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_vline(xintercept = 0, linetype = "dashed") +
  facet_wrap(~ rel_label, scales = "free") +
  scale_colour_manual(
    values = c("PO/FS > 0" = "#1b9e77", "PO/FS <= 0" = "#7570b3"),
    name = "PO vs. FS support"
  ) +
  xlab("log likelihood ratio: parent-offspring minus unrelated") +
  ylab("log likelihood ratio: parent-offspring minus full sibling") +
  theme_bw()
```

For a compact summary, we keep the top-ranked candidate parents for each
offspring. In the larger development workflow these tables were written to disk
for downstream ROC analyses. In a vignette it is clearer to print the relevant
table directly.

```{r top-candidate-parents}
top_pairs <- all_pairs %>%
  arrange(kid_id, desc(logl_ratio)) %>%
  group_by(kid_id) %>%
  slice_head(n = 5) %>%
  ungroup() %>%
  left_join(
    pars_in_samples,
    by = join_by(kid_id == ped_id),
    relationship = "many-to-one"
  )

top_pairs %>%
  select(kid_id, par_id, logl_ratio, po_fs_logl_ratio, rel_label,
         dom_relat, max_hit,
         ped_p1, ped_p2, num_parents_in_sample,
         nonMissingDiag, nonMissingVar)
```

The key columns in this final table are `kid_id`, `par_id`, `logl_ratio`, and
`po_fs_logl_ratio`. The `rel_label`, `dom_relat`, `max_hit`, `ped_p1`, and
`ped_p2` columns come from the simulation and are included only to help verify
the example. In an empirical analysis, candidate pairs with large
log-likelihood ratios would be carried forward for whatever downstream
filtering, ranking, or manual review is appropriate for the study design.
